What We Eval in the Shadows

### 2021-04-14 paper/main.tex
envir

The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. We propose the
following high-level classification the environments of \eval:

\begin{compactitem}[---]
\item \emph{Function:} these environments correspond to some function active on
  the call stack and are obtained by functions such as \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} these are environments that built from data structures
  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
  \c{list2env} or \c{as.environment}. It also includes the empty environment.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environments of a loaded library.
\end{compactitem}

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
function environment, with \c{global} as a distant second. This means that most
variable lookups and most side effects will take place amongst the local
variables of some function. But which function? Table~\ref{tab:funoffset} shows the offset from
the function that directly called \eval, thus 0 is the direct caller and 1 is
its parent. The data suggests that over 80\% of function environments are the
direct caller -- this means, from a user's point of view -- the environment of
the function where \eval occurs textually. It is interesting that some 1.5\% of
sites evaluate a code in an environment that is three frame or above from the
caller of \eval. This distance makes it more tricky to reason about the behavior
of code, as the actions of \eval may happen at a distance.

@@

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
   \begin{tabular}{@{}c|c|c@{}} \hline
       \#kinds & \#sites &  \%sites \\ \hline
       \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
       \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
       \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
       \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\
       \hline
   \end{tabular} \label{tab:polyenvir}
   \caption{Multiplicities}
\end{minipage}\hspace{-1cm}
\end{table}

>>>>

The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environment of \eval can be
classified into the following four kinds:

\begin{compactitem}[---]
\item \emph{Function:} environment for the local variables of some function
  currently active on the call stack. Obtained by calling \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} environments built from data structures such as lists,
  dataframes, or constructed explicitly with \c{new.env}, \c{list2env} or
  \c{as.environment}. Also includes the empty environment.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environment of a loaded library.
\end{compactitem}

\noindent
As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and 1 is its parent, and
so on. The data suggests that in 81\% of cases, \eval access its caller's
environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frame or above from \eval. This
distance implies that, in general, it is modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance.

@@

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
 \begin{tabular}{@{}c|c|c@{}} \hline
 \#kinds & \#sites &  \%sites \\ \hline
 \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
 \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
 \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
 \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
\end{minipage}\hspace{-1cm}
\end{table}
